### 2019-05-25 chapter_00_open_science.tex
Correct till end of chapter opensci

We shall now discuss these topics in greater detail. The chapter is organized
into three sections. The first section focusses on the technical aspects, with
particular focus on emerging developments in programming languages, tools and
services suited for open-source software development. In the next section, we
present Python's strengths and weaknesses as a language for scientific
computing.
%
The section ends with a note on the capabilities of FluidDyn
packages\footnote{We use FluidDyn (with capital letters) to name the project
and \fluiddyn for the base package.}.
%
The last section compares various software development methodologies. Here, we
will also discuss about the possible contradiction between productivity of
individuals and productivity at the community level.
%
% The chapter finally ends with a summary of the motivations for the project
% FluidDyn and an overview of the packages used.

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline, have flourished a lot in the latter half of the twentieth century.
Incidentally the origin of this discipline can be traced back to the ideas in
\citet{richardson_weather_1922} to achieve numerical weather prediction.
Following the arrival of first electronic general-purpose computer, ENIAC, the
first simulations were made using a simple barotropic model to make 24 hour
forecasts \citep{lynch_richardson_2010}. In those early days when the
availability of computers was limited, such methods were neither widespread nor
feasible. Scientific investigations were dominated by theoretical and
experimental methods.
%
We have come a long way since then and today computational methods now
complement theoretical and experimental studies, and are equally important.
%
% Thus, in a typical scientific study, there is significant effort involved in
% translating the mathematical representation to a working research software.
%
With the connectivity provided by the world wide web it is possible to achieve
more and reinvent the way we perform sciences.

@@

In recent years there has been a move towards the open science
\citep[see the report by][]{royal_society_great_britain_science_2012}.
%
Depicted in figure \ref{fig:opensci} are some of steps needed to ensure
reproducibility and transparency in sciences.
%
Traditionally, \emph{knowledge} has been disseminated in the form of courses
and books from universities, and to the wider scientific community through
research articles, conference proceedings and workshops.
%
In the last decade, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals\footnote{In the European
  Union, a proposal to implement Plan S (\url{https://www.coalition-s.org}) was
  been set in motion in September 2018 to fast track adoption of open-access.
} either through systematic
reforms in academia on how researchers' productivity are assessed or through
regulations and offering incentives, such as waiving the application processing
charge for publishing \citep{nosek_promoting_2015}.

A lot of work is required to condense and implement this knowledge in the
form of working applications.
%
The \emph{know-how} can be made accessible and open to scrutiny by making
research software open-source with an appropriate license\footnote{For a
  detailed comparison of licenses, visit either
  \url{https://choosealicense.com} or
  \url{https://www.gnu.org/licenses/license-list.html}.
}.
%
To make codes usable they have to be complemented by documentation which
typically includes tutorials, examples and a detailed commentary of different
components.
%
Furthermore, to inspire confidence in and to ensure reliability of a code
throughout the process of software development, it needs unit-tests through
which a code is continuously built, tested and deployed --- a process known as
Continuous Integration (CI), which is described in detail later in this
chapter. Depending on how matured a software is CI can be easy or
cumbersome to implement, but nevertheless, the result is a cleaner and reliable
code.

Finally, the \emph{results} generated from the such codes together with the
workflow (i.e. scripts to run the codes and to postprocess the data) when made
open in the form of datasets can ensure that research is reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

Open-source software development that we know today has been built on top of
the firm foundations of the Free and Open-Source Software (FOSS) movement. The
term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success for scientific publication
of the free type-setting standards \TeX\ (1977) and \LaTeX\ (1985).
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system.
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development in \citet{raymond_cathedral_1999}, wherein a code can stay
structured despite releasing frequently and delegating tasks to community with
myriad agendas.

@@

Over the years, FOSS development has transitioned from an organic community of
volunteers, towards an organized system with participation from industries,
non-profit organizations and government institutions.
%
Such a model results in a win-win scenario -- users benefit from transparency and
ability to tinker, and the organizations profit with more contributions.
%
The FOSS movement has thus entered into a second era
\citep{fitzgerald_transformation_2006}.

@@

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial). In web-based source
development platforms like GitHub, Bitbucket and GitLab DVCS repositories are
enhanced with code-review tools, which typically consists of a issue tracker
and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and lowering the learning
barrier for scientists who are not experts in software development is crucial
for our project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established based
packages (\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

@@

\subsubsection{Foreign functions}
%
Python is primarily a ``glue'' language, and a common solution to achieve
performance is to use existing code written in C, C++ or Fortran.
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Similarly \href{https://cppyy.readthedocs.io}{cppyy} can interpret C++ code.
Libraries with Python bindings can also be used directly (for e.g., wrapped
using SWIG or Boost.Python) or one has to write extensions, i.e.\ Python
modules written in C or C++ using the CPython API that can be imported directly
from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning
curve compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the scientific
Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

@@

Since fluid dynamics is an engineering science, closed-source commercial software
have historically been dominant in the field.
%
Few people are aware of the vicious circle of the closed-source model for people
and groups relying on it.
%
This mechanism is nevertheless very damaging. In the closed-source model, a group
pays for a license or a new development. The group does not learn how to develop
what has been paid for. The company, on contrary, stands to gain from profits and
the feedback generated from users, through which it improves the product sold.
The group produces codes, books, courses using the closed-source product or
acquire knowledge on how to use the closed-source product, so that the group is
increasingly dependent on the product and is ready to pay more for it.
%
Such vicious circles are difficult to break, but not impossible. This can be
done by introducing alternative open-source solutions, bringing in positivity to
this circle.

We have seen that it was very common to rely on commercial companies and
closed-source software for most of the new developments in fluid mechanics.
%
Nevertheless there are also a lot of codes produced by scientists in the
field. But without the technical knowledge on how to work collectively on code,
this production is often of bad quality (compared to the information technology
industry standard) and it is basically doomed to be abandoned after its use.
%
However, it is natural to try to reuse code or at least to build the next steps
on what has been done before.  Thousands of hours of highly qualified people
are spent in trying to understand and reuse codes badly written with
inappropriate languages!

On a technical point of view, the majority of the coding in the field involve a
mix of Fortran/C or C++, shell languages (as Bash) and Matlab. For experiments,
the graphical programming environment
\href{http://www.ni.com/en-us/shop/labview.html}{Labview} is dominant for
control of physical objects and acquisition and Matlab is used for data
processing.
%
Often, languages are used for things for which they are not adapted.
%
Fortran, C or C++ have not been designed for fast prototyping of complex
programs.
%
It is known that for scientific purposes, one should restrict shell scripting
to extremely simple tasks.
%
Similarly, using Matlab for developing complex programs is technically a
questionable choice.
%
These technical solutions also represent a problem for code reuse, sharing and
collaborative development.
%
We are now going to present factual arguments explaining why we think that the
massive usage of Matlab decreases the collective efficiency of the community.

@@

% see http://www.pyzo.org/python_vs_matlab.html
% see https://www.linkedin.com/pulse/matlab-vs-python-jan-rhebergen

Matlab is an example of a closed-source proprietary numerical computing
environment.
%
It is a good tool for simple processing with matrices, image processing, data
visualization and certain niche applications. The language is well adapted for
these tasks. The development environment is nice and the interpreter is quite
fast, especially now that it has a JIT compiler.

However, the language suffers from serious technical issues which make Matlab
an inadequate tool
% \citationneeded[opinion] % pa: I think what we say is fine for an article. We
% are not writing a wikipedia page which has the ambition to be neutral. Here,
% we gives an argued opinion, which is totally normal in a scientific article.
% av: fair enough :)
for doing more than simple processing and data plotting. The comparison with
Python often debated upon. We list a few striking Matlab weaknesses:

@@

\item No real organization of the standard library. All built-in and
user-defined functions are available in a huge flat namespace. No import
mechanism and as a result we cannot deduce, simply by looking at the code,
where a function comes from.

\item A standard way to organize multi-file code is to write scripts that
modify and define global variables (see for example
\href{https://www.damtp.cam.ac.uk/user/jrt51/files/diablo\_mat.tar.gz}{%
this Matlab version of the code Diablo}).  It is so simple to do this that we
can assert the language strongly encourages this practice.  Matlab files are
not self consistent, i.e.\ it is normal to use in a file a global variable
defined outside of the file.

\item Very bad default argument mechanism.

Default arguments for a function is a very common feature is many programming
languages. In Python, we can write:
\begin{minted}[fontsize=\footnotesize]{python}
def myfunc(a, b, c=1, has_to_print=True):
    if has_to_print:
        print('a =', a, 'b =', b, 'c =', c)
    return c * (a + b)
\end{minted}

An implementation for approximately the same behavior in Matlab could be (as
advised in the official Matlab documentation):
\begin{minted}[fontsize=\footnotesize]{matlab}
function ret = myfunc(a, b, varargin)
    if nargin < 2 | nargin > 4
        error(['The number of arguments has to be ' ...
               'greater than 2 and lower than 5'])
    end
    if nargin == 4
        has_to_print = varargin{2};
    else
        has_to_print = 1;
    end
    if nargin >= 3
        c = varargin{1};
    else
        c = 1;
    end

@@

\item Less intuitive syntax for string operations in Matlab, compared to
Python. Take for example for string comparison, \codeinline{strcmp(s1, s2)} in
Matlab versus \codeinline{s1 == s2} in Python or for look up,
\codeinline{contains(s1, pattern)} in Matlab versus \codeinline{pattern in s1}
in Python.
% \item Less intuitive and clear syntax for exception
% handling. \citationneeded[opinionated].
% pa: Matlab has now a MException class. The syntax is not nice, but it is
% not so important.
\item Less clear and intuitive model and syntax for object oriented
programming, compared to Python. As for functions, one file per user-defined
class is necessary.
% \citationneeded[opinionated].
\item Parenthesis used for both function calls and indexing, causing ambiguity.
\item Matlab codes are usually full of \codeinline{;}, \codeinline{.*},
\codeinline{./}, \codeinline{\&\&} and \codeinline{||}, which make them quite
``noisy''.  Moreover, there is nothing like the Python PEP~8. This makes a
large proportion of the Matlab codes difficult to read and understand, due to
lack of a consistent code style.
\item The syntax \codeinline{a(100, 100) = 1;} to create and extend matrix is a
very good way to hide bugs. The code \codeinline{a = eye(2); a(i0, i1) = 1;}
will never raise any error regardless the values of \codeinline{i0} and
\codeinline{i1}! No error is raised if a user misspell the variable
\codeinline{a} and write something like \codeinline{aa(100, 100) = 1;}.
\end{itemize}

Now we turn our attention, from structural issues of Matlab syntax, to more
practical constraints its use brings about.  By definition of a closed-source
software, Matlab is a big black box: it is impossible to study the
implementation of its functions.

Matlab is not free. The price for one license for non-commercial use is not
very expensive for most research and teaching institutes. However, when we
start to add up the price of the toolboxes necessary to run most codes, it
starts to become substantially large for some institutions.  The price of
Matlab quickly escalates when deployed on a cluster with MPI, since a licence
is required per user, per node!
% av: not per process.
% see: https://www.nas.nasa.gov/hecc/support/kb/running-matlab-applications-in-parallel_522.html

It would be interesting to estimate what could be done with open-source
solutions with a part of the amount paid by public research institutes. How
many developers could be paid to improve the alternative open-source solutions?
%
This question raises the issue of funding scientific open-source software.

Moreover, it is well known that the lack of license for the personal computers
of staff is sometimes solved with cracked versions of Matlab. How can serious
research projects rely on such illegal methods?
%
Note that open-source Matlab interpreters do exist, such as Octave and Scilab.
%
% pa: warning! Octave has a JIT compiler!
%
% Unfortunately, these are not serious alternative today. Their user
% community is too small and they are not very efficient, especially now that
% Matlab uses a JIT compiler and that optimized Matlab codes rely on it.

The recent open-source and Python dynamics represent a hard punch for Matlab,
which is losing the dominant position it has in some domains.
%
Note that this is a very recent development. The Python scientific ecosystem can
now compete in nearly all domains, but this was not the case typically 5 years
ago.
%
There are still a large number of researchers and teachers in sciences who are
strongly attached to Matlab mainly because they have used it a lot.
%
It is, of course, difficult to change habits especially when those are efficient
(for a short time-span and on an individual or group scale) enough for what you
do.

It is normal that Mathworks, the company which makes money with Matlab, spreads
\fnref{https://www.mathworks.com/products/matlab/matlab-vs-python.html}{arguments
against Python and its scientific ecosystem}.  Some of them are valid, for
example the issue of the lively, huge and thus complicated scientific ecosystem
that we have already discussed.  Other arguments are unfair or no longer valid
nowadays.
% Here, we remove a long argument with Mathworks (see the file
% old_arguments_mathworks.tex)
% av: Less provocative, that is good. But we can still add a footnote linking
% the Pyzo page / LinkedIn page comparing python and matlab.
%
% We replace by this:
The scientific Python ecosystem is now very mature, robust, well documented and
fully integrated.  Many of this components are supported by professional
developers paid by companies or research institutes, for example \pack{Numba}
(Anaconda, Nvidia), \pack{TensorFlow} (Google), \pack{PyTorch} (Facebook, Twitter,
Nvidia), Scikit-learn (INRIA), Mercurial (Facebook).
%
The support by the Python community is both very reactive and of very good
level (for example in Stack Overflow, and issues page on major repositories)
and it is now possible to pay companies to get professional support on
scientific Python (for example
\href{https://software.intel.com/en-us/distribution-for-python/get-help}{Intel}
and \href{https://www.anaconda.com/enterprise/}{Anaconda}).
%
Finally, with the tools available today, with little effort we can get a very
efficient program using only Python code.

To conclude on Matlab, some people may use it efficiently for their work, but it
starts to be a problem for the collective efficiency of the community. Except
from the existence of a large code base difficult to translate in better
languages or for some very specific applications, it seems difficult today to
find a good technical reason to continue using Matlab.

The same remarks can be made for Labview. It is a graphical programming
language and the programs are saved as binary files. This is a problem because
first it is impossible to read a Labview program without Labview and, more
importantly, it forbids the use of source management tools which are so
important for collective work and open-source.

Matlab and Labview also lead to problems for students, who learn bad coding
habits and languages that are much less in demand by employers than for example
Python.

For symbolic computation, some specialists use Mathematica, which is a good and
very expensive proprietary closed-source mathematical symbolic computation
program.
%
Sage and Sympy are two complementary open-source alternative both based on
Python.

\paragraphbf{Different models for software development in fluid mechanics}

>>>>

We shall now discuss these topics in greater detail. The chapter is organized
into four sections. The first section focusses on the technical aspects, with
particular focus on emerging developments in programming languages, tools and
services suited for open-source software development. The subsequent section
presents Python's strengths and weaknesses as a language for scientific
computing.
%
The third section compares various software development methodologies. Here, we
will also discuss about the possible contradiction between productivity of
individuals and productivity at the community level.
%
The chapter finally ends with a summary of the motivations for the project
FluidDyn and an overview of the packages we have developed.
% The chapter ends with a note on the capabilities of FluidDyn
% packages\footnote{We use FluidDyn (with capital letters) to name the project
% and \fluiddyn for the base package.}.
%

@@

Computational sciences, and especially computational fluid dynamics (CFD) as a
discipline, have flourished a lot in the latter half of the twentieth century.
Incidentally the origin of this discipline can be traced back to the ideas in
\citet{richardson_weather_1922} to achieve numerical weather prediction.
Following the arrival of first electronic general-purpose computer, ENIAC, the
first simulations were made using a simple barotropic model to make 24 hour
forecasts \citep{lynch_richardson_2010}. In those early days when the
availability of computers was limited, such methods were neither widespread nor
feasible. Scientific investigations were dominated by theoretical and
experimental methods.
%
We have come a long way since then and today computational methods complement
theoretical and experimental studies, and are equally important.
%
% Thus, in a typical scientific study, there is significant effort involved in
% translating the mathematical representation to a working research software.
%
With the connectivity provided by the world wide web it is possible to achieve
more and reinvent the way we do sciences.

@@

In recent years, there has been a move towards open science
\citep[see the report by][]{royal_society_great_britain_science_2012}.
%
Depicted in figure~\ref{fig:opensci} are some of steps needed to ensure
reproducibility and transparency in sciences.
%
Traditionally, \emph{knowledge} has been disseminated in the form of courses
and books from universities, and to the wider scientific community through
research articles, conference proceedings and workshops.
%
In the last decade, knowledge has become more accessible to the public
through public-domain and Creative Commons (or similar) licensed course
materials and massively open online courses (MOOCs). Moreover, there is also an
emerging push to publish in open-access journals\footnote{In the European
  Union, a proposal to implement Plan S (\url{https://www.coalition-s.org}) was
  been set in motion in September 2018 to fast track adoption of open-access.
} either through systematic
reforms in academia on how researchers' productivity are assessed or through
regulations and offering incentives, such as waiving the application processing
charge for publishing \citep{nosek_promoting_2015}.

A lot of work is required to condense and implement this knowledge in the form
of working applications.
%
The \emph{know-how} can be made accessible and open to scrutiny by making
research software open-source with an appropriate license\footnote{For a
	detailed comparison of licenses, visit either
	\url{https://choosealicense.com} or
	\url{https://www.gnu.org/licenses/license-list.html}. }.
%
To make codes usable they have to be complemented by documentation which
typically includes tutorials, examples and a detailed commentary of different
components.
%
Furthermore, to inspire confidence in and to ensure reliability of a code
throughout the process of software development, it needs unit-tests through
which a code is continuously built, tested and deployed --- a process known as
Continuous Integration (CI). Depending on how matured a software is, CI can be
easy or cumbersome to implement, but nevertheless, the result is a cleaner and
reliable code.

Finally, the \emph{results} generated from the such codes together with the
workflow (i.e.\ scripts to run the codes and to post-process the data) when made
open in the form of datasets can ensure that research is reproducible and
preserved in archives for several years to come \citep{gewin_data_2016}. Open
datasets can be shared and made citable through services such as
\href{https://zenodo.org}{Zenodo} and \href{https://figshare.com}{Figshare}.

@@

Open-source software development that we know today has been built on top of
the firm foundations of the Free and Open-Source Software (FOSS) movement. The
term \emph{free} in FOSS is a misnomer, as it actually stands for
\emph{freedom} (to use, modify and distribute). The FOSS movement has
dramatically decreased the cost of using computers, which is evident from the
widespread use of GNU/Linux systems in desktops, computing clusters and
web-servers in academia and beyond.
%
The FOSS culture can be traced back to the success for scientific publication
of the free type-setting standards \TeX\ (1977) and \LaTeX\ (1985).
%
Another founding moment for the FOSS movement was the launch of the GNU project
in 1983 by \href{https://en.wikipedia.org/wiki/Richard_Stallman}{Richard
  Stallman}, to create a Unix-like computer operating system composed entirely
  of free software\footnote{For completeness, see also the work done on
  \href{https://www.levenez.com/unix/}{other Unix operating systems, for
example BSD}.}.
%
GNU is today known for its compiler collection (GCC) and a multitude of tools
which combined with the Linux kernel (created by
\href{https://en.wikipedia.org/wiki/Linus_Torvalds}{Linus Torvalds} in 1991)
forms the GNU/Linux operating system
% and its several distributions (Debian, Red Hat, Slackware, Arch Linux and
% Gentoo to name a few)
that we are familiar with today.
%
% FOSS movement has had huge successes in many frontiers --- Apache could be
% termed as the ``first killer-app of Linux'' and now with an ever-increasing
% suite of softwares including Firefox, LibreOffice, Gimp as solid alternatives
% to proprietary offerings.
%
Linux has become the most widely used kernel, being deployed on servers,
personal computers, embedded devices, and also smart phones (with Android). The
success of the Linux kernel project was attributed to the \emph{bazaar} model of
software development, as described in \citet{raymond_cathedral_1999}, wherein a
code can stay structured despite releasing frequently and delegating tasks to
community with myriad agendas.

@@

Over the years, FOSS development has transitioned from an organic community of
volunteers, towards an organized system with participation from industries,
non-profit organizations and government institutions.
%
Such a model results in a win-win scenario -- users benefit from transparency and
ability to tinker, and the organizations profit with more contributions.
%
The FOSS movement has now entered into a second era
\citep{fitzgerald_transformation_2006}.

@@

Collaboration was achieved in the early years of FOSS development through
emailing ``patches'', and centralized Version Control Systems (VCS) near the
turn of the 21\textsuperscript{st} century.  The collective collaboration on
development is today streamlined with the emergence of Distributed Version
Control Systems (DVCS), especially, Git and Mercurial). In web-based source
development platforms like GitHub, Bitbucket and GitLab, DVCS repositories are
enhanced with code-review tools, typically consisting of a issue tracker
and pull-request mechanism.
%
Git has become the \textit{de facto} standard for source management. However,
for the FluidDyn project, we prefer to use Mercurial because we consider it to
be simpler for the beginners and just as powerful for expert users\footnote{See
  \href{https://www.mercurial-scm.org/wiki/GitConcepts}{``a detailed
comparison''} on Mercurial website.}. Simplicity and lowering the learning
barrier for scientists who are not experts in software development is crucial
for our project.
%
Note that, it is now possible to use Mercurial to work with Git repositories,
for example hosted in GitHub and GitLab\footnote{By using the Mercurial
extension \href{http://hg-git.github.io/}{hg-git}.}.
%
Version control tools are designed to work on any kind of text files, not just
codes, so they could even be utilized to write scientific documents with
multiple authors.
%
The use of DVCS in science will lead to long term productivity benefits
\citep{wilson_best_2014}.

@@

Python
% has recently become one of the most popular language for teaching programming
% and developing scientific applications. It 
has now a mature and powerful scientific ecosystem with well-established based
packages (\Numpy for N-dimensional homogeneous arrays, \Scipy as the
fundamental toolkit for scientific computing, \pack{Matplotlib} for plotting
and \pack{Pandas} for data structures) and several more specialized packages
(to name a few, \pack{h5py}, \pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras},
\pack{TensorFlow} and \pack{PyTorch}.
%
There user-friendly and ready-to-use Python installers (similar to Matlab), in
particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda} to streamline the installation in 
desktop platforms.
Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{https://pyzo.org}{Pyzo},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab} are often useful for
interactive development.

@@

\subsubsection{Foreign functions}
%
Python is primarily a ``glue'' language, and a common solution to achieve
performance is to use existing code written in C, C++ or Fortran.
%
It is possible, without a compilation step, to call functions in C libraries
through pure python code using \pack{ctypes} standard module or the \pack{cffi}
package.
%
Similarly \href{https://cppyy.readthedocs.io}{\pack{cppyy}} can interpret C++
code. Libraries with Python bindings can also be used directly (for e.g.,
wrapped using SWIG or Boost.Python) or one has to write extensions, i.e.\
Python modules written in C or C++ using the CPython API that can be imported
directly from Python like any Python module.
%
However, to rely on such techniques to call C libraries would inhibit
development of extending functionalities, as they have a steeper learning curve
compared to Python.
%
Nevertheless, this strategy has given rise to the base modules of the
scientific Python stack, namely \Numpy, \Scipy and \pack{Matplotlib}.

@@

Since fluid dynamics is an engineering science, closed-source commercial software
have historically been dominant in the field.
%
Few people are aware of the vicious circle of the closed-source model for people
and groups relying on it.
%
This mechanism is nevertheless very damaging. In the closed-source model, a group
pays for a license or a new development. The group does not learn how to develop
what has been paid for. The company, on the contrary, gains from
the feedback generated from users, through which it improves the product.
The group produces codes, books, courses using the closed-source product or
acquire knowledge on how to use the closed-source product, making the group
increasingly dependent on the product and ready to pay more for it.
%
It is difficult to break such vicious circles, but not impossible. This can be
done by introducing alternative open-source solutions.

Despite the reliance on commercial software for research in fluid mechanics,
%
there are also a lot of codes produced by scientists in fluid mechanics. But
without the technical knowledge on how to work collectively, such codes
are often of bad quality (compared to the information technology
industry standard) and are basically doomed to be abandoned after their use.
%
However, it is natural to try to reuse codes or at least to build the next steps
on what has been done already. Thousands of hours of highly qualified people
are spent in trying to understand and reuse badly written codes in
inappropriate languages!

From a technical perspective, the majority of the coding in the field involves
a mix of Fortran/C or C++, shell languages (as Bash) and Matlab. For
experiments, the graphical programming environment
\href{http://www.ni.com/en-us/shop/labview.html}{Labview} is dominant for
control of physical objects and acquisition. Fortran is often used to implement
solvers and post-processing algorithms. Matlab is used for data processing and
visualization.
%
Often, languages are used for things for which they are not adapted.
%
Fortran, C or C++ have not been designed for fast prototyping of complex
programs.
%
It is known that for scientific purposes, one should restrict shell scripting
to extremely simple tasks.
%
Use of compiled languages to develop new algorithms hinders the development
process and ability to debug.
%
Similarly, using Matlab for developing complex programs is a questionable
choice.
%
These technical solutions are also problematic for code reuse, sharing and
collaborative development.
%
We are now going to present factual arguments explaining why we think that the
massive usage of Matlab decreases the collective efficiency of the community.

@@

Matlab is an example of a closed-source proprietary numerical computing
environment.
%
It is a good tool for simple processing with matrices, image processing, data
visualization and certain niche applications. The language is well adapted for
these tasks. The development environment is nice and the interpreter is quite
fast, especially now that it has a JIT compiler.
%
However, the language suffers from serious technical issues which make Matlab
an inadequate tool
% \citationneeded[opinion] % pa: I think what we say is fine for an article. We
% are not writing a wikipedia page which has the ambition to be neutral. Here,
% we gives an argued opinion, which is totally normal in a scientific article.
% av: fair enough :)
for doing more than simple processing and data plotting. The comparison with
Python is often debated upon . We list a few striking Matlab weaknesses:

@@

\item No real organization of the standard library. All built-in and
user-defined functions are available in a huge flat namespace. Due to this
absence of an import mechanism we cannot deduce, simply by looking at the code,
where a function is defined.

\item A standard way to organize multi-file code is to write scripts that
modify and define global variables (see for example
\fnref{https://www.damtp.cam.ac.uk/user/jrt51/files/diablo\_mat.tar.gz}{%
the Matlab version of the code Diablo}).  It is so simple to do this that we
can assert the language strongly encourages this practice.  Matlab files are
not self consistent, i.e.\ it is normal to use in a file a global variable
defined outside of the file.

\item Very bad default argument mechanism.
%
Default arguments for a function is a very common feature is many programming
languages. In Python, we can write:
\begin{minted}[fontsize=\footnotesize]{python}
def myfunc(a, b, c=1, has_to_print=True):
    if has_to_print:
        print('a =', a, 'b =', b, 'c =', c)
    return c * (a + b)
\end{minted}
%
An implementation for approximately the same behaviour in Matlab could be (as
advised in the official Matlab documentation):
\begin{minted}[fontsize=\footnotesize]{matlab}
function ret = myfunc(a, b, varargin)
    if nargin < 2 | nargin > 4
        error(['The number of arguments has to be ' ...
               'greater than 2 and lower than 5'])
    end
    if nargin == 4
        has_to_print = varargin{2};
    else
        has_to_print = 1;
    end
    if nargin >= 3
        c = varargin{1};
    else
        c = 1;
    end

@@

\item Non-intuitive syntax for string operations in Matlab, compared to
Python. Take for example for string comparison, \codeinline{strcmp(s1, s2)} in
Matlab versus \codeinline{s1 == s2} in Python or for look up,
\codeinline{contains(s1, pattern)} in Matlab versus \codeinline{pattern in s1}
in Python.
\item Complicated syntax for object oriented programming, compared to Python.
As for functions, one file per user-defined class is necessary.
\item Parentheses used for both function calls and indexing, causing ambiguity.
\item Matlab codes are usually full of \codeinline{;}, \codeinline{.*},
\codeinline{./}, \codeinline{\&\&} and \codeinline{||}, which make them quite
``noisy''.  Moreover, due to a lack of a consistent code style like, Python's
PEP~8, a
large proportion of the Matlab codes are difficult to read and understand.
\item The syntax \codeinline{a(100, 100) = 1;} to create and extend a matrix is a
very good way to hide bugs. The code \codeinline{a = eye(2); a(i0, i1) = 1;}
will never raise any error regardless the values of \codeinline{i0} and
\codeinline{i1}! No error is raised if a user misspell the variable
\codeinline{a} and write something like \codeinline{aa(100, 100) = 1;}.
\end{itemize}
%
Now we turn our attention, from structural issues of Matlab syntax, to more
practical constraints brings about. The arguments are equally applicable for
any proprietary software used in research.  Closed-source
software are, by definition, black boxes: it is impossible to study the
implementation of its functions.
%
Matlab is not free. The price for one license for non-commercial use is not
very expensive for most research and teaching institutes. However, when we
start to add up the price of the toolboxes necessary to run most codes, it
starts to become substantially large for some institutions.  The price of
Matlab quickly escalates when deployed on a cluster with MPI, since a licence
is \fnref{%
	https://www.nas.nasa.gov/hecc/support/kb/running-matlab-applications-in-parallel_522.html}{%
required per user, per node}.
%
Moreover, it is well known that the lack of license for the personal computers
of staff are solved with pirated versions. How can serious
research projects rely on such illegal methods?
%
It is also unfair in education when students have to work on course projects and
assignments, with a Matlab code as a starting point and not a free alternative.

The same remarks can be made for Labview. It is a graphical programming
language and the programs are saved as binary files. As a result, it is
impossible to read a Labview program without Labview and to apply version
control, crippling collaborative possibilities.
%
Matlab and Labview also lead to problems for students, who learn bad coding
habits and languages that are much less in demand by employers than, for example
\fnref{https://insights.stackoverflow.com/survey/2018\#technology}{C++ or
Python}.

Note that open-source Matlab interpreters do exist, such as Octave and Scilab.
%
With some minimal effort, one can port existing Matlab codes using either
Julia which by design uses Matlab-like syntax, or Python's \codeinline{pylab}
package provided via \pack{matplotlib} and the command-line tool
\codeinline{fluidmat2py} in \pack{fluiddyn}.
%
For symbolic computation, some specialists use Maple or Mathematica, which are
good but expensive proprietary closed-source programs.
%
Sage and Sympy are two complementary open-source alternative both based on
Python.
%
The \fluidpack{lab} project in FluidDyn project is an attempt to orchestrate
experiments without relying on Labview.

It is normal that Mathworks, the company which makes money with Matlab, spreads
\fnref{https://www.mathworks.com/products/matlab/matlab-vs-python.html}{arguments
against Python and its scientific ecosystem}.  Some of them are valid, for
example the issue of the lively, huge and thus complicated scientific ecosystem
that we have already discussed. Other arguments are unfair or no longer valid
nowadays\footnote{See also \url{http://www.pyzo.org/python\_vs\_matlab.html}
or \url{https://www.linkedin.com/pulse/matlab-vs-python-jan-rhebergen}}.
%
The Python scientific ecosystem can now compete in nearly all domains, but this
was not the case typically 5 years ago.
%
It would be interesting to estimate what could be done with open-source
solutions with a part of the amount paid by public research institutes. How
many developers could be paid to improve the alternative open-source solutions?
%
This question raises the issue of funding scientific open-source software.

\subsection{Different models for software development in fluid mechanics}
